Brain region specific differential expression analysis (all 7 regions/one region at a time)

### 2017-05-14 code/Rmd/regionLevelDiffExpAnlz.Rmd
Added AAD differential expression function to region level differential expression analysis

```{r fxn}
performDifferentialExpression <- function(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc){

  diffExp = foreach(i = 1:n, 
                    .combine = rbind,
                    .packages = c('limma','edgeR','cqn','dplyr','plyr','CovariateAnalysis','data.table'),
                    .export = c('gene.len', 'gene.gc')) %dopar% {

                    # Sub-sample data for balanced comparison
                    minSampSz = min(summary(cov[,contVar]), na.rm = T)
                    innerCov = rownameToFirstColumn(cov, 'SampleID')  %>%
                      plyr::ddply(.variables = contVar, .fun = function(x, minSampSz){
                        ind = sample(1:dim(x)[1], minSampSz)
                        return(x[ind,])
                      }, minSampSz) %>%
                      droplevels()
                    rownames(innerCov) = innerCov$SampleID

                    # Get design matrix
                    designMat = getDesignMatrix(innerCov[,varOfInterest], Intercept = F)$design
                    designMat = designMat[,linColumnFinder(designMat)$indepCols]

                    # Compute offset for gene length and gc content
                    cqn.gene.expression = cqn(cnt[,innerCov$SampleID],
                                              x = gene.gc[rownames(cnt), 'percentage_gc_content'],
                                              lengths = gene.len[rownames(cnt), 'gene.length'],
                                              lengthMethod = "smooth",
                                              verbose = FALSE)
                    cqn.gene.expression$E = cqn.gene.expression$y+cqn.gene.expression$offset

@@

                    # Estimate voom weights for dispersion control
                    voom.weights = voom(cnt[,innerCov$SampleID], design = designMat, plot=F, na.rm = T)

@@

                    # Perform differential expression
                    contrasts = levels(cov[,contVar]) %>%
                      paste(contVar,.,sep = '') %>%
                      combn(2) %>%
                      apply(2,paste, collapse = '-') %>%
                      makeContrasts(contrasts = ., levels = colnames(baseModel$coefficients))
                    baseModel.cntr =  contrasts.fit(baseModel, contrasts)
                    baseModel.cntr = eBayes(baseModel.cntr)
                    innerDiffExp = lapply(1:dim(contrasts)[2], function(i, baseModel.cntr){
                      topTable(baseModel.cntr, coef = i, number = 50000, confint = T) %>%
                        rownameToFirstColumn('ensembl_gene_id')
                    }, baseModel.cntr)
                    names(innerDiffExp) = colnames(contrasts)
                    innerDiffExp = innerDiffExp %>%
                      rbindlist(use.names = T, fill = T, idcol = 'Comparison') %>%
                      dplyr::mutate(iter = i,
                                    Comparison = gsub(contVar,'',Comparison))

                    return(innerDiffExp)
                    }

  diffExp = diffExp %>%
    dplyr::group_by(Comparison, ensembl_gene_id) %>%
    tidyr::gather(feature, value, -Comparison, -ensembl_gene_id) %>%
    dplyr::group_by(Comparison, ensembl_gene_id, feature) %>%
    dplyr::summarise(value = median(value, na.rm = T)) %>%
    tidyr::spread(feature, value)

@@

### Perform differential expression (per brain region)
```{r diff.exp}
all.results = plyr::dlply(covariates, .(BrainRegion), .fun = function(cov, cnt, gene.len, gene.gc){
  n = 100
  # Filter counts and covariates
  cnt = cnt[,cov$SampleID]

@@

  baseDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc)

@@

  genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, n, gene.len, gene.gc)

  return(rbindlist(list(baseDiffExp, genderDiffExp), use.names = T, fill = T))
}, counts, gene.len, gene.gc)

@@

writeLines('Number of differentially expressed genes between CONTROL.FEMALE-CONTROL.MALE at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'CONTROL.FEMALE-CONTROL.MALE', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(CONTROL.FEMALE_vs_CONTROL.MALE = DF) %>%
  kable(row.names=F)
```

>>>>

```{r fxn}
performDifferentialExpression <- function(cov, cnt, contVar, varOfInterest, gene.len, gene.gc){

  # Get design matrix
  designMat = getDesignMatrix(cov[,varOfInterest], Intercept = F)$design
  designMat = designMat[,linColumnFinder(designMat)$indepCols]

  # Compute offset for gene length and gc content
  cqn.gene.expression = cqn(cnt,
                            x = gene.gc[rownames(cnt), 'percentage_gc_content'],
                            lengths = gene.len[rownames(cnt), 'gene.length'],
                            lengthMethod = "smooth",
                            verbose = FALSE)
  cqn.gene.expression$E = cqn.gene.expression$y+cqn.gene.expression$offset

@@

  # Estimate voom weights for dispersion control
  voom.weights = voom(cnt, design = designMat, plot=F, na.rm = T)

@@

  # Perform differential expression
  contrasts = levels(cov[,contVar]) %>%
    paste(contVar,.,sep = '') %>%
    combn(2) %>%
    apply(2,paste, collapse = '-') %>%
    makeContrasts(contrasts = ., levels = colnames(baseModel$coefficients))
  baseModel.cntr =  contrasts.fit(baseModel, contrasts)
  baseModel.cntr = eBayes(baseModel.cntr)
  diffExp = lapply(1:dim(contrasts)[2], function(i, baseModel.cntr){
    topTable(baseModel.cntr, coef = i, number = 50000, confint = T) %>%
      rownameToFirstColumn('ensembl_gene_id')
    }, baseModel.cntr)
  names(diffExp) = colnames(contrasts)
  diffExp = diffExp %>%
    rbindlist(use.names = T, fill = T, idcol = 'Comparison') %>%
    dplyr::mutate(Comparison = gsub(contVar,'',Comparison))

  return(diffExp)
}

performDifferentialExpressionAAD <- function(cov, cnt, contVar, varOfInterest, gene.len, gene.gc){

  # Get design matrix
  designMat = getDesignMatrix(cov[,varOfInterest], Intercept = F)$design
  designMat[,grep('Diagnosis',colnames(designMat))] = designMat[,grep('Diagnosis',colnames(designMat))] * designMat[,'AAD']
  colnames(designMat)[grep('Diagnosis',colnames(designMat))] = paste(grep('Diagnosis',colnames(designMat), value = T), 'AAD', sep = '.')
  designMat = designMat[,setdiff(colnames(designMat), 'AAD')]
  designMat = designMat[,linColumnFinder(designMat)$indepCols]

  # Compute offset for gene length and gc content
  cqn.gene.expression = cqn(cnt,
                            x = gene.gc[rownames(cnt), 'percentage_gc_content'],
                            lengths = gene.len[rownames(cnt), 'gene.length'],
                            lengthMethod = "smooth",
                            verbose = FALSE)
  cqn.gene.expression$E = cqn.gene.expression$y+cqn.gene.expression$offset

  # Estimate voom weights for dispersion control
  voom.weights = voom(cnt, design = designMat, plot=F, na.rm = T)

  # Perform differential expression
  contrasts = levels(cov[,contVar]) %>%
    paste(contVar,.,sep = '') %>%
    paste('AAD',sep = '.') %>%
    combn(2) %>%
    apply(2,paste, collapse = '-') %>%
    c('DiagnosisAD.AAD', 'DiagnosisCONTROL.AAD') %>%
    makeContrasts(contrasts = ., levels = colnames(baseModel$coefficients))
  baseModel.cntr =  contrasts.fit(baseModel, contrasts)
  baseModel.cntr = eBayes(baseModel.cntr)
  diffExp = lapply(1:dim(contrasts)[2], function(i, baseModel.cntr){
    topTable(baseModel.cntr, coef = i, number = 50000, confint = T) %>%
      rownameToFirstColumn('ensembl_gene_id')
    }, baseModel.cntr)
  names(diffExp) = colnames(contrasts)
  diffExp = diffExp %>%
    rbindlist(use.names = T, fill = T, idcol = 'Comparison') %>%
    dplyr::mutate(Comparison = gsub(contVar,'',Comparison))

@@

### Perform differential expression (per brain region)
```{r diff.exp}
all.results = plyr::dlply(covariates, .(BrainRegion), .fun = function(cov, cnt, gene.len, gene.gc){

  # Filter counts and covariates
  cnt = cnt[,cov$SampleID]

@@

  baseDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)

@@

  genderDiffExp <- performDifferentialExpression(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)

  # Perform differential expression between case-control in a AAD specific manner
  varOfInterest = c('Diagnosis', 'Gender', 'Batch', 'RIN', 'AAD', 'PMI', 
                    'PCT_PF_READS_ALIGNED', 'PCT_INTERGENIC_BASES', 'PCT_INTRONIC_BASES', 
                    'PCT_RIBOSOMAL_BASES')
  contVar = 'Diagnosis'

  aadDiffExp <- performDifferentialExpressionAAD(cov, cnt, contVar, varOfInterest, gene.len, gene.gc)

  return(rbindlist(list(baseDiffExp, genderDiffExp, aadDiffExp), use.names = T, fill = T))
}, counts, gene.len, gene.gc)

@@

writeLines('Number of differentially expressed genes between CONTROL.FEMALE-CONTROL.MALE at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'CONTROL.FEMALE-CONTROL.MALE', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(CONTROL.FEMALE_vs_CONTROL.MALE = DF) %>%
  kable(row.names=F)

writeLines('Number of differentially expressed genes between AD.AAD-CONTROL.AAD at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'AD.AAD-CONTROL.AAD', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(AD.AAD_vs_CONTROL.AAD = DF) %>%
  kable(row.names=F)

writeLines('Number of differentially expressed genes between AD.AAD at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'AD.AAD', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(AD.AAD = DF) %>%
  kable(row.names=F)

writeLines('Number of differentially expressed genes between CONTROL.AAD at an FDR of 0.05 and fold change of 1.2 are')
sapply(all.results, function(x){
  sum(x$adj.P.Val <= 0.05 & abs(x$logFC) >= log2(1.2) & x$Comparison == 'CONTROL.AAD', na.rm = T)
}) %>% rownameToFirstColumn('Region') %>% dplyr::rename(CONTROL.AAD = DF) %>%
  kable(row.names=F)
```
